---
title: "conseguiR: A Lightweight Object-First Walkthrough"
author: "The conseguiR authors"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{conseguiR: A Lightweight Object-First Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r libraries}
library(conseguiR)
library(data.table)
library(igraph)
```

# Overview

`conseguiR` is designed as an object-first workflow for integrating:

- germline evidence
- somatic evidence
- epigenomic evidence

onto a gene-regulatory graph.

In a full analysis, those three evidence streams are combined, imposed onto a
backend graph, propagated by diffusion, and then distilled into smaller
interpretable outputs such as score plots, locus plots, and selected gene-gene
subgraphs.

This vignette deliberately uses a **tiny in-memory toy example** rather than
large real study files. The goal here is to show how the package behaves as an
R package:

- compute stages return objects
- plots can be generated from those objects
- external tools such as MAGMA and Python-backed diffusion are additional
  runtime stages, not the only way to understand the package API

# The object-first idea

The highest-level package design is:

1. validate inputs
2. compute modality-specific score objects
3. build a scored graph object
4. optionally run downstream diffusion and subgraph calling
5. visualize intermediate or downstream objects

The important package behavior is that intermediate compute stages return
bundles in memory by default. Users do not need to create a pile of files just
to move from one stage to the next.

# A tiny validation example

We start with minimal in-memory GWAS-like and somatic inputs. This is enough to
demonstrate the lightweight validation layer.

```{r validation-inputs}
toy_gwas <- data.table(
  variant_id = c("rs1", "rs2", "rs3"),
  chromosome = c("1", "1", "1"),
  base_pair_location = c(101L, 220L, 340L),
  p_value = c(0.05, 1e-4, 0.2)
)

toy_somatic <- data.table(
  sample_id = c("S1", "S2", "S2"),
  chromosome = c("1", "1", "1"),
  start_position = c(105L, 230L, 360L),
  end_position = c(105L, 230L, 360L),
  ref = c("A", "G", "C"),
  alt = c("T", "A", "T")
)

validation <- validate_inputs(
  gwas_sumstats = toy_gwas,
  somatic_maf = toy_somatic,
  verbose = FALSE
)

validation$bundle_type
```

This answers a modest but important question: are the supplied objects
structured well enough for the package to start working with them?

# Building a tiny scored graph

For the rest of the vignette we create a tiny toy gene-regulatory graph in
memory. This avoids heavy external dependencies while still exercising the
package's graph-scoring interface.

```{r toy-graph}
toy_vertices <- data.frame(
  name = c("MYC", "BCL2", "GH01J000001", "GH01J000002"),
  node_id = c("MYC", "BCL2", "GH01J000001", "GH01J000002"),
  node_type = c("gene", "gene", "reg", "reg"),
  stringsAsFactors = FALSE
)

toy_edges <- data.frame(
  from = c("GH01J000001", "GH01J000001", "GH01J000002"),
  to = c("MYC", "BCL2", "MYC"),
  confidence = c(0.9, 0.6, 0.8),
  stringsAsFactors = FALSE
)

toy_graph <- graph_from_data_frame(
  d = toy_edges,
  vertices = toy_vertices,
  directed = TRUE
)
```

Now we create tiny modality-specific score tables. These are not meant to be
biologically meaningful; they are just enough to show the API contract.

```{r toy-scores}
toy_gene_germline <- data.table(
  gene_id = c("MYC", "BCL2"),
  zstat = c(3.2, 1.4)
)

toy_reg_germline <- data.table(
  reg_elem_id = c("GH01J000001", "GH01J000002"),
  zstat = c(2.6, 1.1)
)

toy_gene_somatic <- data.table(
  gene_id = c("MYC", "BCL2"),
  zstat = c(1.8, -0.9),
  p_value = c(0.02, 0.4)
)

toy_reg_somatic <- data.table(
  reg_elem_id = c("GH01J000001", "GH01J000002"),
  zstat = c(0.7, -0.4),
  p_value = c(0.3, 0.7)
)

toy_reg_epigenomic <- data.table(
  reg_elem_id = c("GH01J000001", "GH01J000002"),
  zstat = c(2.1, 0.3)
)
```

The scored-graph stage accepts either full score bundles from earlier package
stages or explicit score tables. Here we use the explicit table route.

```{r scored-graph}
scored_graph <- build_scored_gene_reg_graph(
  graph = toy_graph,
  gene_germline_scores = toy_gene_germline,
  reg_germline_scores = toy_reg_germline,
  gene_somatic_scores = toy_gene_somatic,
  reg_somatic_scores = toy_reg_somatic,
  reg_epigenomic_scores = toy_reg_epigenomic,
  save_outputs = FALSE,
  verbose = FALSE
)

scored_graph$bundle_type
```

The returned bundle contains the scored graph object plus the scored node and
edge tables.

```{r scored-nodes}
scored_graph$objects$nodes[, .(
  node_id,
  node_type,
  germline_score,
  somatic_score,
  epigenomic_score
)]
```

This is the core pre-diffusion integration step:

- genes receive germline and somatic scores
- regulatory elements receive germline, somatic, and epigenomic scores
- missing modalities are filled with zeros rather than silently dropping nodes

# Plotting score objects

Because plotting functions accept bundles and tables directly, users can inspect
intermediate results without first saving anything to disk.

```{r germline-plot}
germline_gene_plot <- plot_germline_gene_scores(
  scored_graph = scored_graph,
  stage = "pre",
  label_features = c("MYC", "BCL2"),
  save_plot = FALSE
)

germline_gene_plot$plot
```

We can do the same for the regulatory epigenomic scores.

```{r epigenomic-plot}
epigenomic_reg_plot <- plot_epigenomic_reg_scores(
  scored_graph = scored_graph,
  label_features = c("GH01J000001"),
  save_plot = FALSE
)

epigenomic_reg_plot$plot
```

This demonstrates an important package pattern: plotting can happen straight
from in-memory bundle objects, and file output is optional.

# Runtime-aware stages

Some package stages depend on additional external runtimes:

- MAGMA for germline scoring from real GWAS data
- managed Python via `basilisk` for diffusion and selected-subgraph calling

The runtime check helper tells us whether those heavier stages are currently
available.

```{r runtime-check}
runtime_status <- check_conseguiR_runtime(quiet = TRUE)

list(
  core_ok = runtime_status$core_ok,
  python_stage_ok = runtime_status$python_stage_ok,
  magma_ok = runtime_status$magma_ok
)
```

That lets users decide whether they can proceed directly into the full pipeline
or whether they first need to install MAGMA or allow the Python-backed stages
to provision.

# Full pipeline shape

In a real analysis, the intended high-level sequence is still:

```{r full-pipeline-shape, eval=FALSE}
germline_scores <- prepare_germline_scores(...)
somatic_scores <- prepare_somatic_scores(...)
epigenomic_scores <- prepare_epigenomic_scores(...)

scored_graph <- build_scored_gene_reg_graph(
  germline_scores = germline_scores,
  somatic_scores = somatic_scores,
  epigenomic_scores = epigenomic_scores
)

diffusion <- run_gene_reg_diffusion(scored_graph = scored_graph)
selected_subgraph <- call_selected_subgraph(diffusion = diffusion)
```

Those stages are not run in this lightweight vignette because they depend on
external tools and real study-scale inputs. The toy example above focuses on
what the package itself contributes most directly:

- object-oriented handoff between stages
- graph-based score integration
- bundle-driven plotting

# Conclusion

This toy walkthrough is intentionally small, but it captures the core package
philosophy:

- `conseguiR` works with in-memory objects, not just files
- graph scoring is a first-class stage in the workflow
- visualization functions can inspect intermediate objects directly
- heavier external-runtime stages are layered on top of that object-first base

That makes the package easier to debug, easier to script, and easier to extend
to real germline, somatic, and epigenomic analyses once the relevant runtimes
and study files are available.

```{r session-info}
sessionInfo()
```
