---
title: "conseguiR Worked Example: SLE, Lymphoid Mutations, and DLBCL Epigenomics"
author: "The conseguiR authors"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: false
    number_sections: true
    latex_engine: xelatex
vignette: >
  %\VignetteIndexEntry{conseguiR Worked Example: SLE, Lymphoid Mutations, and DLBCL Epigenomics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "92%"
)
```

```{r libraries}
library(conseguiR)
library(data.table)
library(knitr)
```

<style>
body {
  line-height: 1.75;
  font-size: 16px;
}
h1, h2, h3 {
  font-weight: 700;
}
.lead {
  font-size: 1.1em;
  color: #4b5563;
}
.callout {
  padding: 1rem 1.2rem;
  border-left: 4px solid #2c7fb8;
  background: #f7fbff;
  margin: 1.2rem 0;
}
</style>

# Executive summary

<div class="lead">
This report presents a worked <code>conseguiR</code> analysis built around a
practical three-modality question: how do germline risk, somatic mutation
enrichment, and regulatory variability combine on a shared gene-regulatory
graph? For the vignette we use an SLE GWAS, a PCAWG lymphoid-malignancy MAF
with CLL excluded, and a deliberately small three-track DLBCL H3K27ac panel.
The motivating biological hypothesis is that <code>NAMPT</code> may represent a
shared metabolic vulnerability linking autoimmune dysregulation and lymphoma
biology. We use that hypothesis as a focused narrative thread rather than as a
claim that the vignette alone definitively proves.
</div>

At a high level, the workflow proceeds in six stages:

1. validate the raw inputs
2. compute germline, somatic, and epigenomic scores
3. impose those scores on the gene-regulatory graph
4. run diffusion to obtain post-diffusion gene scores
5. extract a smaller selected subgraph
6. inspect the results through score plots, network plots, and a locus plot

Because some compute stages are intentionally heavy, the code below is shown as
runnable analysis code but is not evaluated during document render. The figures
embedded later in the report are generated outputs saved from the worked
analysis path, so the document still reads like an analysis report rather than
just an API listing.

# Example inputs

The example analysis uses the following files.

```{r inputs}
gwas_path <- "data/raw/Testing/GCST90436637.h.tsv"
somatic_maf_path <- "data/raw/Testing/2026-01-09_no_CLL_lymph_only_pcawg_maf_tcga_order_hg38.maf"
reg_ref_path <- "data/processed/GRCh38-cCREs.loc"
refdb_path <- "data/raw/Testing/RefCDS_human_GRCh38.p12.rda"
reference_bfile <- "data/raw/g1000_eur/g1000_eur"
aid_covariate_path <- "data/raw/Testing/2026-01-26_all_reg_elems_sample_level_mut_frac_comparison_bet_only_memory_b_normal_and_non_cll_malig_b_cells.rds"
sample_size <- 399917L

bw_files <- c(
  "data/raw/Testing/SRR1020516_DLBCL_P286_H3K27ac_ChIPseq.bw",
  "data/raw/Testing/SRR1554599_DLBCL_DHL6_H3K27ac_ChIPseq.bw",
  "data/raw/Testing/SRR7758629_DLBCL_U2932_H3K27ac_ChIPseq.bw"
)

data.table(
  object = c(
    "GWAS",
    "Somatic MAF",
    "Regulatory reference",
    "dndscv reference",
    "PLINK reference",
    "fishHook covariate data",
    "Selected bigWig tracks"
  ),
  value = c(
    gwas_path,
    somatic_maf_path,
    reg_ref_path,
    refdb_path,
    reference_bfile,
    aid_covariate_path,
    paste(basename(bw_files), collapse = ", ")
  )
)
```

<div class="callout">
For the vignette we intentionally keep the epigenomic branch small. The three
DLBCL tracks above were chosen to give a clean, lymphoma-centered example while
still making the cross-track variability step easy to understand.
</div>

# Step 1: validate the inputs

Before running any heavy computation, we usually make sure the basic file and
column structure is sound.

```{r validate-code, eval=FALSE}
validation <- validate_inputs(
  gwas_sumstats = gwas_path,
  somatic_maf = somatic_maf_path,
  reg_ref_path = reg_ref_path,
  epigenomic_tracks = bw_files,
  verbose = TRUE
)
```

The validation stage is intentionally lightweight. It is meant to catch:

- missing files
- missing key columns
- unreadable regulatory references
- unreadable bigWig headers

It is not meant to fully dry-run MAGMA, dndscv, fishHook, or diffusion.

# Step 2: compute the three score modalities

## Germline scoring with MAGMA

The germline branch uses MAGMA twice, once for genes and once for regulatory
elements. In this worked example the input GWAS is an SLE summary-statistics
table, so the germline layer represents inherited autoimmune risk rather than a
lymphoma-only signal.

```{r germline-code, eval=FALSE}
germline_scores <- prepare_germline_scores(
  gwas_sumstats = gwas_path,
  reference_bfile = reference_bfile,
  gene_sample_size = sample_size,
  reg_sample_size = sample_size,
  gene_step1_args = list(annotation_window = c(35, 10)),
  gene_step2_args = list(
    gene_model = "snp-wise=mean",
    pval_use = c("SNP", "P"),
    pval_duplicate = "drop"
  ),
  reg_step1_args = list(annotation_window = c(0, 0)),
  reg_step2_args = list(
    gene_model = "snp-wise=mean",
    pval_use = c("SNP", "P"),
    pval_duplicate = "drop"
  ),
  verbose = TRUE
)
```

The key MAGMA idea is:

- step 1 decides which SNPs belong to each feature
- step 2 decides how SNP-level signal becomes one feature-level score

## Somatic scoring with dndscv and fishHook

The somatic branch is split into a gene model and a regulatory model.

In the current package defaults, the gene branch asks `dndscv` for one-sided
positive and negative selection output when that API surface is available, and
the regulatory branch uses fishHook's directional `p` and `p.neg` columns when
present. That keeps score extraction aligned with the underlying statistical
tests rather than reconstructing direction from a generic two-sided p-value.

```{r somatic-code, eval=FALSE}
covariate_dt <- readRDS(aid_covariate_path)

somatic_scores <- prepare_somatic_scores(
  maf = somatic_maf_path,
  refdb = refdb_path,
  reg_ref_path = reg_ref_path,
  dndscv_args = list(sm = "192r_3w"),
  fishhook_covariate_data = covariate_dt,
  fishhook_args = list(),
  verbose = TRUE
)
```

For this vignette the fishHook covariate is a precomputed **AID-per-base**
background measure supplied in a ready-to-use table. Biologically, it
represents expected local exposure to activation-induced cytidine deaminase
(AID) mutagenesis. That matters in B-cell malignancies because some regulatory
elements accumulate mutations partly because they sit in an AID-exposed
background, not necessarily because they are under stronger positive selection.

Using a precomputed covariate keeps the vignette light: fishHook receives a
table that is already aligned to the regulatory-element universe instead of
recomputing sequence or sample-level summaries during document execution.

The practical table shape is still simple:

```{r fishhook-example, eval=FALSE}
head(covariate_dt[, c("reg_elem_id", "n_samples_mut_norm")])

covariate_dt <- data.frame(
  reg_elem_id = c("EH38E0080197", "EH38E2084302"),
  n_samples_mut_norm = c(3.8, 0.9)
)
```

That is, one row per regulatory element, a regulatory-element identifier, and
the covariate values fishHook should model. Inside `conseguiR`, that table is
converted into a fishHook-ready AID-per-base covariate after matching each
regulatory element back onto the ENCODE cCRE universe.

## Epigenomic scoring from bigWig tracks

The epigenomic score in `conseguiR` is best interpreted as a **cross-track
variability score**, not a generic activity score. For the vignette we keep
this branch intentionally small and DLBCL-centered by using three H3K27ac
tracks:

- `SRR1020516_DLBCL_P286_H3K27ac_ChIPseq.bw`
- `SRR1554599_DLBCL_DHL6_H3K27ac_ChIPseq.bw`
- `SRR7758629_DLBCL_U2932_H3K27ac_ChIPseq.bw`

This makes the example easier to follow than a much larger track panel while
still preserving the idea that regulatory elements are being ranked by how much
their activity changes across related malignant contexts.

```{r epigenomic-code, eval=FALSE}
epigenomic_scores <- prepare_epigenomic_scores(
  reg_ref_path = reg_ref_path,
  bw_files = bw_files,
  min_tracks = 3L,
  drop_mhc = TRUE,
  transform = "log1p",
  summary_fun = mean,
  verbose = TRUE
)
```

# Step 3: build the scored gene-regulatory graph

```{r graph-code, eval=FALSE}
scored_graph <- build_scored_gene_reg_graph(
  germline_scores = germline_scores,
  somatic_scores = somatic_scores,
  epigenomic_scores = epigenomic_scores,
  verbose = TRUE
)
```

At this point, all three modalities have been imposed onto the graph. Missing
modalities are zero-filled by default, which keeps the graph broad enough to
retain biologically meaningful features even when one branch is absent.

# Step 4: diffusion

As in the overview vignette, the diffusion and selected-subgraph stages are
Python-backed internally but are run through conseguiR's managed `basilisk`
environment rather than a user-supplied interpreter.

```{r diffusion-code, eval=FALSE}
diffusion <- run_gene_reg_diffusion(
  scored_graph = scored_graph,
  top_k = 3L,
  confidence_power = 2.0,
  beta_germline = 0.5,
  beta_somatic = 0.5,
  beta_epigenomic = 0.7,
  integration_weight_germline = 1.0,
  integration_weight_somatic = 1.0,
  integration_weight_epigenomic = 1.0,
  reg_signal_clip = 5.0,
  verbose = TRUE
)
```

This stage turns the pre-diffusion graph scores into a gene-centered
post-diffusion ranking that incorporates regulatory context.

The main ranking columns now use a weighted signed Stouffer-style integration
across the three post-diffusion modalities. That means positive germline,
somatic, and epigenomic support reinforce one another, while negative modality
contributions are carried forward and can reduce the final integrated score.

# Step 5: selected subgraph calling

```{r subgraph-code, eval=FALSE}
selected_subgraph <- call_selected_subgraph(
  diffusion = diffusion,
  target_genes = 50L,
  candidate_pool_size = 400L,
  node_prize_weight = 1,
  edge_conf_weight = 1,
  edge_cost_weight = 1,
  verbose = TRUE
)
```

This is the step where the workflow shifts from a full graph to a smaller,
more interpretable network.

# Step 6: figures and interpretation

In a full analysis these plotting calls would be evaluated and saved to disk.
For the package vignette we keep the code visible, and then show the generated
figure artifact immediately after each call.

## Germline gene plot

```{r code-germline-plot, eval=FALSE}
plot_germline_gene_scores(
  germline_scores = germline_scores,
  stage = "pre",
  label_features = c("MYC", "BCL2", "CXCR5"),
  plot_file_path = "germline_genes_pre_rank.pdf"
)
```

This plot gives us a quick read on which genes are leading the germline signal
before diffusion.

```{r fig-germline, echo=FALSE, out.width='95%'}
knitr::include_graphics("../data/processed/figures/germline_genes_pre_rank.pdf")
```

## Somatic gene plot

```{r code-somatic-gene-plot, eval=FALSE}
plot_somatic_gene_scores(
  somatic_scores = somatic_scores,
  stage = "pre",
  label_features = c("MYC", "BCL2", "TP53"),
  plot_file_path = "somatic_genes_pre_volcano.pdf"
)
```

This is where we can see which genes dominate the somatic branch before
regulatory context is allowed to propagate by diffusion.

```{r fig-somatic-gene, echo=FALSE, out.width='95%'}
knitr::include_graphics("../data/processed/figures/somatic_genes_pre_volcano.pdf")
```

## Epigenomic regulatory plot

```{r code-epi-plot, eval=FALSE}
plot_epigenomic_reg_scores(
  epigenomic_scores = epigenomic_scores,
  label_features = c("NAMPT"),
  plot_file_path = "epigenomic_regs.pdf"
)
```

This figure emphasizes the regulatory elements that are most variable across
the selected DLBCL epigenomic tracks. In this vignette, that epigenomic branch
is meant to support a focused lymphoma narrative rather than to exhaustively
profile every available bigWig.

```{r fig-epi-reg, echo=FALSE, out.width='95%'}
knitr::include_graphics("../data/processed/figures/epigenomic_regs.pdf")
```

## Selected subgraph plot

```{r code-subgraph-plot, eval=FALSE}
plot_selected_subgraph(
  selected_subgraph = selected_subgraph,
  plot_file_path = "selected_subgraph.pdf"
)
```

This is usually the point where the workflow becomes easiest to discuss with
other people, because we have moved from long ranked tables to a smaller and
more interpretable network view.

```{r fig-subgraph, echo=FALSE, out.width='95%'}
knitr::include_graphics("../data/processed/figures/gene_gene_selected_subgraph_plot.pdf")
```

## Locus-level plot

```{r code-locus-plot, eval=FALSE}
plot_locus_context(
  chromosome = "7",
  start = 106200000,
  end = 106330000,
  scored_graph = scored_graph,
  diffusion = diffusion,
  gwas_sumstats = gwas_path,
  label_top_lit_snps = 3L,
  label_features = c("NAMPT"),
  plot_file_path = "SLE_NAMPT_locus_context.pdf",
  verbose = TRUE
)
```

The locus plot is where the package becomes especially useful for biological
storytelling. In this worked example, it gives us a direct way to inspect the
`NAMPT` neighborhood while keeping the three input modalities in the same view.
It lets us connect:

- regulatory-element scores
- post-diffusion gene scores
- regulatory interactions
- and cited SNPs that overlap the locus

```{r fig-nampt-locus, echo=FALSE, out.width='98%'}
knitr::include_graphics("../data/processed/figures/NAMPT_locus_context.pdf")
```

In that sense, `NAMPT` functions as a scientifically plausible shared
metabolic-vulnerability candidate in the report: the germline branch asks
whether autoimmune risk accumulates in the region, the somatic branch asks
whether lymphoid mutations are enriched beyond background expectation, and the
epigenomic branch asks whether the linked regulatory landscape is variable
across malignant DLBCL states.

# Closing thoughts

This report is meant to read like an analysis narrative rather than a bare API
listing. The important pattern across the workflow is that each stage adds a
different layer of structure:

- input validation establishes trust in the raw data
- scoring provides modality-specific signal
- graph construction puts all modalities in one shared space
- diffusion makes the regulatory context gene-centered
- subgraph calling distills the large network into something more interpretable
- plotting turns those results into views that can actually support discussion

In practice, that is why `conseguiR` works best as a staged workflow rather
than as a one-line black box.

# Session information

```{r session-info}
sessionInfo()
```
